Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS

### 2011-07-20 ms.tex
Edited citations to match

\altaffiltext{1}{Corresponding authors: N. Karlsson, nkarlsson@physics.umn.edu \& C. Pfrommer, christoph.pfrommer@h-its.org}
\altaffiltext{2}{School of Physics and Astronomy, University of Minnesota, Minneapolis, MN 55455, USA}
\altaffiltext{3}{DESY, Platanenallee 6, 15738 Zeuthen, Germany}
\altaffiltext{4}{ Institut fu\"r Physik und Astronomie, Universit\"at Potsdam, 14476 Potsdam-Golm, Germany}
\altaffiltext{5}{Heidelberg Institute	for Theoretical	Studies, Schloss-Wolfsbrunnenweg 35, D-69118 Heidelberg, Germany}
\altaffiltext{6}{Canadian Institute for Theoretical Astrophysics, University of Toronto, Toronto, ON, M5S 3H8, Canada}
\altaffiltext{7}{University of California, Santa Barbara, CA 93106}

@@

\begin{abstract}
Theoretical models of galaxy clusters suggest they may be sources of very high-energy (VHE) gamma-ray emission. Observations of radio halos and relics in clusters indicate populations of relativistic electrons and positrons, which are either injected into the intra-cluster medium and accelerated by turbulent shocks or secondaries from interactions of cosmic-ray hadrons (protons and ions) with the intra-cluster gas. Those interactions would also produce gamma rays through the decay of neutral pions. Reported here are observations of the Coma cluster of galaxies with the VERITAS array of imaging Cherenkov telescopes. The observations were made between March and May 2008. Radio observations at 1.41 GHz taken with the Green Bank Telescope and X-ray observations with the ROSAT All-Sky survey was used to a priori select regions of interest in the field of view. No significant VHE gamma-ray emission from the Coma cluster was detected by VERITAS. Several emission regions are considered, including the cluster core as both unresolved (point-like) and spatially extended, and a select few cluster members. The measured flux upper limits at the 99\% confidence level are on the order of $2-5\times 10^{-8}$ ph. m$^{-2}$ s$^{-1}$, which is X times higher than the model-predicted minimum gamma-ray flux.

@@

%\keywords{cosmic rays: interactions --- galaxies --- galaxy clusters -- VHE}
\keywords{cosmic rays --- gamma rays: observations --- gamma rays: VHE --- galaxies: clusters: general --- galaxies: clusters: individual: Coma (ACO 1656) --- dark matter}

@@

The Coma cluster of galaxies (ACO 1656) is one of the most thouroughly studied clusters over all wavelengths \citep{article:Voges_etal:1999}. Located at a distance of about 100 Mpc \citep[$z=0.023$;][]{article:StrubleRood:1999}, it is one of the closest and also one of the most massive clusters \citep[$M \sim 10^{15}M_{\odot}$;][]{article:Smith:1983, article:Kubo_etal:2008}. It hosts both a giant radio halo \citep{article:Giovannini_etal:1993,article:Thierbach_etal:2003} and peripheral radio relic, which appears connected to the radio halo with a ``diffuse'' bridge \citep[see discussion in][]{article:BrownRudnick:2010}. It has been suggested that the relic is an infall shock \citep{1998A&A...332..395E} and succesively demonstrated by cosmological simulations that model the non-thermal emission processes \citep{2008MNRAS.385.1211P,2008MNRAS.385.1242P,2009MNRAS.393.1073B, 2011ApJ...735...96S}. Extended soft X-ray (SXR) emission is evident from the ROSAT all-sky survey in the 0.1 to 2.4 keV band \citep{article:BrielHenryBohringer:1992} and observations with XMM-Newton \citep{article:Briel_etal:2001} revealed substructure in the X-ray halo. As such it is a natural candidate for gamma-ray observations. In this article, results from the VERITAS observations of the Coma cluster of galaxies are reported, together with the resulting constraints on cosmic-ray particle populations, magnetic fields, and dark matter in the cluster. Throughout the analysis, we use a present day Hubble constant of $H_0 = 100 h \mbox{ km s}^{-1} \mbox{ Mpc}^{-1}$ with $h=0.7$.

@@

\section{Fermi-LAT Observations and Analysis}
The Large Area Telescope (LAT) on board the \emph{Fermi} Gamma-ray Space Telescope (\emph{Fermi}) has observed the Coma cluster in all-sky survey mode since its launch in June 2008. The \emph{Fermi}-LAT is sensitive to gamma rays in the 20 MeV to $\sim$300 GeV energy range and is thus very complimentary to the VERITAS obervations. \citet{article:Ackermann_etal:2010} reported on the search for gamma-ray emission from thirty-three galaxy clusters in the data from the first 18 months, including the Coma cluster, for which an upper limit of $4.58\times 10^{-9}$ ph. cm$^{-2}$ s$^{-1}$ in the 0.2 to 100 GeV energy band was reported. This limit is expected to improve as the exposure is increased. In this work an updated analysis which includes additional data taken during the period 2010 February 5 to 2011 June 10 is presented as a compliment to the VERITAS results.

The analysis of this work follows the same procedure as described in detail in \citet{article:Abdo_etal:2009}. To eliminate albedo gamma rays from the Earth limb, a zenith angle cut of 105$^{\circ}$ is applied. The \emph{Fermi} Science Tools version v9r23p1 and Instrumental Response Functions (IRFs) P6\_V11\_Diffuse (a model of the spatial distribution of photon events calibrated ...) were used throughout this work. Only photons above 1 GeV are considered, where the constraints on the flux are most meaningful. The 99\% confidence level flux upper limits in the energy range 1-100 GeV, assuming both a unresolved, point-like or spatially extended emission are tabulated in Table \ref{tabel:fermi}. The gamma-ray emission is assumed to follow a power-law distribution with a photon index $\alpha=2.1$ and $2.5$.

@@

\section{Cosmic Rays}

We decided to adopt a multifaceted approach to constrain the CR-to-thermal
pressure distribution in Coma with these upper limits. That includes (1) an
analytic model derived from cosmological hydrodynamical simulations of the
formation of galaxy clusters, (2) a simplified multi-frequency analytical
approach that assumes a constant CR-to-thermal energy density and a power-law
spectrum in momentum, and (3) an approach that uses the observed luminosity of
the giant radio halo in Coma to place a lower limit on the expected $\gamma$-ray
flux in the hadronic model -- independent of the magnetic field distribution.
This translates into a minimum CR pressure which, if challenged by tight
gamma-ray limits/detections, enables scrutinizing the hadronic interaction model
of the formation of giant radio halos.  Modelling the CR distribution through
different techniques enable to bracket our ignorance about the underlying plasma
physics that shape the CR distribution and hence reflect the Bayesian priors
that are imposed on the modelling \citep[see][for a
discussion]{2011arXiv1105.3240P}.

\subsection{Simulation-based approach}

We adopt the universal spectral and spatial gamma-ray model developed by
\citet{2010MNRAS.409..449P} to estimate the emission from decaying neutral pions
that dominates over the IC emission from primary and secondary electrons above
100~MeV in clusters. Given a density profile as, e.g., inferred by cosmological
simulations or X-ray observations, the analytic approach models the CR
distribution and the associated radiative emission processes from radio to the
gamma-ray band.  This formalism was derived from high resolution simulations of
clusters of galaxies that included radiative hydrodynamics, star formation,
supernova feedback, and followed CR physics using a novel formulation that trace
the most important injection and loss processes self-consistently while
accounting for the CR pressure in the equation of motion
\citep{2006MNRAS.367..113P,2007A&A...473...41E,2008A&A...481...33J}.  The
results are in line with earlier numerical results on some of the overall
characteristics of the CR distribution and the associated radiative emission
processes \citep{2000A&A...362..151D,2001ApJ...559...59M,2003MNRAS.342.1009M,
  2007MNRAS.378..385P,2008MNRAS.385.1211P,2008MNRAS.385.1242P}, but for the
first time is able to track the spectral evolution as a function of space while
accounting for the non-linear feedback of such a population on the hydrodynamics
in a high-resolution cosmological simulation of the formation of galaxy
clusters.

We note that the overall normalization of the CR and gamma-ray distribution
scales with the maximum acceleration efficiency at structure formation shock
waves. Following recent observations at supernova remnants
\cite{2009Sci...325..719H} as well as theoretical studies
\cite{2005ApJ...620...44K}, we adopt an optimistic but nevertheless realistic
value of this parameter and assume that 50\% of the dissipated energy at strong
shocks is injected into CRs while this efficiency rapidly decreases for weaker
shocks \cite{2007A&A...473...41E}. Since the vast majority of internal formation
shocks (merger and flow shocks) are weak shocks with Mach numbers $M\lesssim3$
\citep[e.g.,][]{2003ApJ...593..599R,2006MNRAS.367..113P} they do not
significantly contribute to the CR population in clusters. Instead, strong
shocks during the formation epoch of clusters and strong accretion shocks at the
present time (at the boundary of voids and filaments/super-cluster regions)
dominate the injection of CRs which are adiabatically transported into clusters
thereafter. Hence, these model predictions provide a plausible upper limit for
the CR contribution from structure formation shocks in galaxy clusters which can
be scaled down for a lower effective acceleration efficiency in the strong shock
regime.  On the other hand, other possible CR sources such as AGN and starburst
driven galactic winds have been neglected for simplicity which could in
principle increase the expected gamma-ray yield from clusters by an unknown
factor.

There are additional limitations of the simulated physics in these cosmological
simulations which only considered advective transport of CRs by turbulent gas
motions which produces centrally enhanced profiles.  However, active CR
transport such as CR diffusion and streaming tends to drive the CR radial
profiles towards being flat, with equal CR number density everywhere. While the
CR streaming is then usually larger than typical advection velocities and
becomes comparable or lower than this only for periods with trans- and
supersonic cluster turbulence during a cluster merger. As a consequence a
bimodality of the CR spatial distribution results with merging (relaxed)
clusters showing a centrally concentrated (flat) CR energy density profile
\cite{2011A&A...527A..99E}.  This translates into a bimodality of the expected
diffuse radio and gamma-ray emission of clusters, since more centrally
concentrated CR will find higher target densities for hadronic CR proton
interactions \cite{2011A&A...527A..99E}. As a result of this, relaxed clusters
could have a reduced gamma-ray luminosity by up to a factor of five. The goal of
this and other future work of seeking for gamma-ray emission from clusters is to
constrain the acceleration physics and transport properties of CRs.

We adopt the electron density profile for the Coma cluster that has been
inferred from ROSAT X-ray observations \citep{1992A&A...259L..31B} and model the
spatial and spectral distribution of CRs according to the (conservative) model
by \citet{2010MNRAS.409..449P} that neglects the contribution of cluster
galaxies. Figure ?? shows the expected spectrum of Coma, normalized to the total
flux within the virial radius. Add {\em Fermi} results \ldots

@@

To complement the simulation-inspired analysis and to facilitate comparison with
earlier analytical work, we repeat the data analysis with a simplified
analytical model that assumes a power-law CR spectrum and a constant relative CR
pressure, i.e., we adopt the {\em isobaric model of CRs}. Since a priori, the CR
spectral index is unconstrained, we vary it within plausible ranges of
$2.1<\alpha_\rmn{p}<2.5$. These spectral ranges are needed to explain the radio
spectral index of the giant radio halo of Coma after accounting for the spectral
steepening at frequencies $\nu\sim5~\rmn{GHz}$ due to the Sunyaev-Zel'dovich
effect \citep{2004A&A...413...17P}\footnote[4]{Assuming a magnetic field of
  1~$\mu$G, the CR protons responsible for the GHz radio emitting electrons have
  an energy of $~100$~GeV and are $\sim$ 20 times less energetic than those CR
  protons that are responsible for the 200 GeV $\gamma$-ray emission. This is
  consistent with the concave curvature found in the universal CR spectrum
  \citet{2010MNRAS.409..449P}.}.  Table~\ref{table:constraints} shows the
resulting constraints on $X_\CR$ which range between 0.05 and 1, strongly
depending on the assumed spectral index.

@@

\del{
the limit for 0.2 deg and alpha_p = 2.3 is compatible with the simulation based
one, X_CR < 0.1. However, there is currently a discrepancy in our assumptions
about the lower energy cutoff of CRs: in the simulations we use P/mp c = 0.8 (as
simulated) while I set this to 0 in my simplified (isobaric) model. The
discrepancy can be understood as follows:
*) in the isobaric model, we would have to normalize our $X_\CR$ by a factor of 2/3
for an effective spectral index of ~ 2.4 (between GeV and 200 GeV) as inferred
from the simulations (see Fig 1, CP & Ensslin 2004, A&A, 413, 17).
*) In the simulation model, we have an increased pressure by this factor of 3/2
(in comparison to a power law spectrum with alpha_TeV ~ 2.3 down to a cutoff in
the normalized momentum of 0.8) due to the concave curvature of our spectrum.
}

@@

For clusters that host radio halos we are able to derive a minimum $\gamma$-ray
flux in the hadronic model of CR interactions. The idea is based on the fact
that a stationary distribution of CR electrons loses all its energy to
synchrotron radiation for strong magnetic fields ($B \gg B_\rmn{CMB} \simeq 3.2
\mu$G) so that the ratio of $\gamma$-ray to synchrotron flux becomes independent
of the spatial distribution of CRs and thermal gas \citep{2008MNRAS.385.1242P}.
This can be easily seen by considering the pion decay induced $\gamma$-ray
luminosity $L_\gamma$ and the synchrotron luminosity $L_\nu$ of a steady state
distribution of CR electrons that has been generated by hadronic CR
interactions,
\begin{eqnarray}
L_\gamma &=& A_\gamma \dps\int d V\, n_\CR n_\rmn{gas},\\
\label{eq:Lnu}
   L_\nu &=& A_\nu \dps\int d V\, n_\CR n_\rmn{gas} \frac{\dps
    \eps_B^{(\alpha_\nu+1)/2}}{\dps\eps_\rmn{CMB}+\eps_B}\\
  &\simeq&A_\nu \dps\int d V\, n_\CR n_\rmn{gas}\quad\rmn{for}~\eps_B \gg \eps_\rmn{CMB}.
\end{eqnarray}
Here $A_\gamma$ and $A_\nu$ are dimensional constants that depend on the
hadronic physics of the interaction \citep{2008MNRAS.385.1211P,
  2008MNRAS.385.1242P} and $\alpha_\nu\simeq 1$ is the observed synchrotron
spectral index. Hence we can derive a minimum $\gamma$-ray flux in the hadronic
model
\begin{equation}
  \label{eq:Fmin}
  \mathcal{F}_{\gamma,\rmn{min}} = 
  \frac{\dps A_\gamma}{\dps A_\nu}\frac{\dps L_\nu}{\dps 4\pi D_\rmn{lum}^2},
\end{equation}
where $L_\nu$ is the observed luminosity of the radio mini-halo and
$D_\rmn{lum}$ denotes the luminosity distance to the respective cluster.
Lowering the magnetic field would require an increase in the energy density of
CR electrons to reproduce the observed synchrotron luminosity and thus increase
the associated $\gamma$-ray flux.

To derive a minimum gamma-ray flux that is comparable to the upper limits, we
need to determined the radio flux within the corresponding angular regions. To
this end, we fit the point source subtracted, azimuthally averaged radio halo
profile at 1.38 GHz \citet{1997A&A...321...55D} with a $\beta$-model,
\begin{equation}
S_\nu (r_\bot)= S_0\, 
\left[ 1 + \left( \frac{r_\bot}{r_{\rmn{c}}}\right)^2\right]
^{-3\beta + 1/2},
\label{beta}
\end{equation}
where $S_0 = 1.1 \times 10^{-3}\,\rmn{Jy\,arcmin}^{-2}$, $r_{\rmn{c}} = 450
kpc$, and $\beta = 0.78$.  Within the error bars, this profile is consistent
with 326 MHz-data taken by \citet{2001A&A...369..441G} when scaled with a radio
spectral index of 1.15.

@@

We have seen in the previous Section that we can obtain an absolute lower limit
on the gamma-ray emission in the hadronic model by assuming high magnetic
fields, $B>B_\rmn{CMB}$. Hence, we can turn this around by using our upper limit
on the gamma-ray emission (and by extension on the CR pressure) to infer a lower
limit on the magnetic field that is needed to explain the observed radio
emission which, however, assumes the validity of the hadronic model. Lowering
the gamma-ray constraint will tighten (increase) the magnetic field limit. Once
this gets in conflict with magnetic field measurements by means of other
methods, e.g., Faraday rotation measure (RM)\footnote{Generally, Faraday RM
  of the magnetic field strength by, e.g., background sources through clusters
  are degenerate with the magnetic coherence scale. Only a simultaneous
  deprojection method that self-consistently solves for the magnetic power
  spectrum is in principle able to break this degeneracy, however, require
  high-quality data and picks up further complications when using embedded
  sources within the cluster plasma, e.g., degeneracies with the assumed
  projection angle of the jet \citep{2011A&A...529A..13K}.}, this would
challenge the hadronic model of radio halos.  The method that we use to
constrain the magnetic field inherits a dependence on the assumed spatial
structure that we parametrize by
\begin{equation}
  \label{eq:B}
  B(r) = B_0  \,\left(\frac{n_\rmn{e}(r)}{n_\rmn{e}(0)}\right)^{\alpha_B},
\end{equation}
as suggested by Faraday RM studies and numerical magneto-hydrodynamical (MHD)
simulations \citep[][and references therein]{2010A&A...513A..30B,
  2011arXiv1107.0968B}. In fact, the magnetic field in the Coma cluster is among
the best ones constrained due to its proximity that enables RM observations of
seven radio-sources located at projected distances of 50 to 1500 kpc from the
cluster center. The best fit model obtains $B_0 = 4.7^{+0.7}_{-0.8}\,\mu$G and
$\alpha_B = 0.5^{+0.2}_{-0.1}$ \citep{2010A&A...513A..30B}.  We aim for
constraining the central field strength, $B_0$, which depends on the magnetic
decline, $\alpha_B$, that we vary (within reasonable ranges of
$\Delta\alpha_B=0.2$ as suggested from the Faraday RM studies) to address this
dependence. We proceed as follows:

>>>>

\altaffiltext{1}{Corresponding authors: N. Karlsson, nkarlsson@physics.umn.edu \& C. Pfrommer, christoph.pfrommer@h-its.org}
\altaffiltext{2}{School of Physics and Astronomy, University of Minnesota, Minneapolis, MN 55455, USA}
\altaffiltext{3}{DESY, Platanenallee 6, 15738 Zeuthen, Germany}
\altaffiltext{4}{ Institut f\"ur Physik und Astronomie, Universit\"at Potsdam, 14476 Potsdam-Golm, Germany}
\altaffiltext{5}{Heidelberg Institute	for Theoretical	Studies, Schloss-Wolfsbrunnenweg 35, D-69118 Heidelberg, Germany}
\altaffiltext{6}{Canadian Institute for Theoretical Astrophysics, University of Toronto, Toronto, ON, M5S 3H8, Canada}
\altaffiltext{7}{Department of Physics, University of California, Santa Barbara, CA 93106, USA}

@@

\begin{abstract}
Theoretical models of galaxy clusters suggest they may be sources of very high-energy (VHE) gamma-ray emission. Observations of radio halos and relics in clusters indicate populations of relativistic electrons, which are either injected into the intra-cluster medium and accelerated by turbulent shocks or secondaries from interactions of cosmic-ray hadrons (protons and ions) with the intra-cluster gas. Those interactions would also produce gamma rays through the decay of neutral pions. Reported here are observations of the Coma cluster of galaxies with the VERITAS array of imaging Cherenkov telescopes. The observations were made between March and May 2008. Radio observations at 1.41 GHz taken with the Green Bank Telescope and X-ray observations with the ROSAT All-Sky survey was used to a priori select regions of interest in the field of view. No significant VHE gamma-ray emission from the Coma cluster was detected by VERITAS. Several emission regions are considered, including the cluster core as both unresolved (point-like) and spatially extended, and a select few cluster members. The flux upper limits at the 99\% confidence level measured are on the order of $2-5\times 10^{-8}$ ph. m$^{-2}$ s$^{-1}$, which is X times higher than the model-predicted minimum gamma-ray flux.

@@

%\keywords{cosmic rays: interactions --- galaxies --- galaxy clusters -- VHE}
\keywords{cosmic rays --- gamma rays: observations --- gamma rays: VHE --- galaxies: clusters: general --- galaxies: clusters: individual: Coma (ACO 1656) --- dark matter --- magnetic fields}

@@

The Coma cluster of galaxies (ACO 1656) is one of the most thouroughly studied clusters over all wavelengths \citep{article:Voges_etal:1999}. Located at a distance of about 100 Mpc \citep[$z=0.023$;][]{article:StrubleRood:1999}, it is one of the closest and also one of the most massive clusters \citep[$M \sim 10^{15}M_{\odot}$;][]{article:Smith:1983, article:Kubo_etal:2008}. It hosts both a giant radio halo \citep{article:Giovannini_etal:1993,article:Thierbach_etal:2003} and peripheral radio relic, which appears connected to the radio halo with a ``diffuse'' bridge \citep[see discussion in][]{article:BrownRudnick:2010}. It has been suggested that the relic is an infall shock \citep{article:Ensslin_etal:1998} and succesively demonstrated by cosmological simulations that model the non-thermal emission processes \citep{article:PfrommerEnsslinSpringel:2008, article:Pfrommer:2008, article:Battaglia_etal:2009, article:Skillman_etal:2011}. Extended soft X-ray (SXR) emission is evident from the ROSAT all-sky survey in the 0.1 to 2.4 keV band \citep{article:BrielHenryBohringer:1992} and observations with XMM-Newton \citep{article:Briel_etal:2001} revealed substructure in the X-ray halo. As such it is a natural candidate for gamma-ray observations. In this article, results from the VERITAS observations of the Coma cluster of galaxies are reported, together with the resulting constraints on cosmic-ray particle populations, magnetic fields, and dark matter in the cluster. Throughout the analysis, a present day Hubble constant of $H_0 = 100 h \mbox{ km s}^{-1} \mbox{ Mpc}^{-1}$ with $h=0.7$ has been used.

@@

\section{Fermi-LAT Observations and Analysis}
The Large Area Telescope (LAT) on board the \emph{Fermi} Gamma-ray Space Telescope (\emph{Fermi}) has observed the Coma cluster in all-sky survey mode since its launch in June 2008. The \emph{Fermi}-LAT is sensitive to gamma rays in the 20 MeV to $\sim$300 GeV energy range and is thus very complimentary to the VERITAS observations. \citet{article:Ackermann_etal:2010} reported on the search for gamma-ray emission from thirty-three galaxy clusters in the data from the first 18 months, including the Coma cluster, for which an upper limit of $4.58\times 10^{-9}$ ph. cm$^{-2}$ s$^{-1}$ in the 0.2 to 100 GeV energy band was reported. This limit is expected to improve as the exposure is increased. In this work an updated analysis which includes additional data taken during the period 2010 February 5 to 2011 June 10 is presented as a compliment to the VERITAS results.

The analysis of this work follows the same procedure as described in detail in \citet{article:Abdo_etal:2009}. To eliminate albedo gamma rays from the Earth limb, a zenith angle cut of 105$^{\circ}$ is applied. The \emph{Fermi} Science Tools version v9r23p1 and Instrumental Response Functions (IRFs) P6\_V11\_Diffuse (a model of the spatial distribution of photon events calibrated  in flight) were used throughout this work. Only photons above 1 GeV are considered, where the constraints on the flux are most meaningful. The 99\% confidence level flux upper limits in the energy range 1-100 GeV, assuming both a unresolved, point-like or spatially extended emission are tabulated in Table \ref{tabel:fermi}. The gamma-ray emission is assumed to follow a power-law distribution with a photon index $\alpha=2.1$ and $2.5$.

@@

\section{Gamma Rays Emission from Cosmic Rays}
We decided to adopt a multifaceted approach to constrain the CR-to-thermal
pressure distribution in Coma with these upper limits. That includes (1) an
analytic model derived from cosmological hydrodynamical simulations of the
formation of galaxy clusters, (2) a simplified multi-frequency analytical
approach that assumes a constant CR-to-thermal energy density and a power-law
spectrum in momentum, and (3) an approach that uses the observed luminosity of
the giant radio halo in Coma to place a lower limit on the expected $\gamma$-ray
flux in the hadronic model -- independent of the magnetic field distribution.
This translates into a minimum CR pressure which, if challenged by tight
gamma-ray limits/detections, enables scrutinizing the hadronic interaction model
of the formation of giant radio halos.  Modelling the CR distribution through
different techniques enable to bracket our ignorance about the underlying plasma
physics that shape the CR distribution and hence reflect the Bayesian priors
that are imposed on the modelling \citep[see][for a
discussion]{article:PinzkePfrommerBergstrom}.

\subsection{Simulation-based approach}
We adopt the universal spectral and spatial gamma-ray model developed by
\citet{article:PinzkePfrommer:2010} to estimate the emission from decaying neutral pions
that dominates over the IC emission from primary and secondary electrons above
100~MeV in clusters. Given a density profile as, e.g., inferred by cosmological
simulations or X-ray observations, the analytic approach models the CR
distribution and the associated radiative emission processes from radio to the
gamma-ray band.  This formalism was derived from high resolution simulations of
clusters of galaxies that included radiative hydrodynamics, star formation,
supernova feedback, and followed CR physics using a novel formulation that trace
the most important injection and loss processes self-consistently while
accounting for the CR pressure in the equation of motion
\citep{article:PfrommerSpringelEnsslinJubelgas, article:EnsslinPfrommerSpringelJubelgas:2007, article:JubelgasSpringelEnsslinPfrommer:2008}.  The
results are in line with earlier numerical results on some of the overall
characteristics of the CR distribution and the associated radiative emission
processes \citep{article:DolagEnsslin, article:MiniatiRyuKangJones:2001, article:Miniati:2003, article:Pfrommer_etal:2007, article:PfrommerEnsslinSpringel:2008, article:Pfrommer:2008}, but for the
first time is able to track the spectral evolution as a function of space while
accounting for the non-linear feedback of such a population on the hydrodynamics
in a high-resolution cosmological simulation of the formation of galaxy
clusters.

We note that the overall normalization of the CR and gamma-ray distribution
scales with the maximum acceleration efficiency at structure formation shock
waves. Following recent observations at supernova remnants
\cite{article:Helder_etal:2009} as well as theoretical studies
\cite{article:KandJones:2005}, we adopt an optimistic but nevertheless realistic
value of this parameter and assume that 50\% of the dissipated energy at strong
shocks is injected into CRs while this efficiency rapidly decreases for weaker
shocks \cite{article:EnsslinPfrommerSpringelJubelgas:2007}. Since the vast majority of internal formation
shocks (merger and flow shocks) are weak shocks with Mach numbers $M\lesssim3$
\citep[e.g.,][]{article:Ryu_etal:2003, article:PfrommerSpringelEnsslinJubelgas} they do not
significantly contribute to the CR population in clusters. Instead, strong
shocks during the formation epoch of clusters and strong accretion shocks at the
present time (at the boundary of voids and filaments/super-cluster regions)
dominate the injection of CRs which are adiabatically transported into clusters
thereafter. Hence, these model predictions provide a plausible upper limit for
the CR contribution from structure formation shocks in galaxy clusters which can
be scaled down for a lower effective acceleration efficiency in the strong shock
regime.  On the other hand, other possible CR sources such as AGN and starburst
driven galactic winds have been neglected for simplicity which could in
principle increase the expected gamma-ray yield from clusters by an unknown
factor.

There are additional limitations of the simulated physics in these cosmological
simulations which only considered advective transport of CRs by turbulent gas
motions which produces centrally enhanced profiles.  However, active CR
transport such as CR diffusion and streaming tends to drive the CR radial
profiles towards being flat, with equal CR number density everywhere. While the
CR streaming is then usually larger than typical advection velocities and
becomes comparable or lower than this only for periods with trans- and
supersonic cluster turbulence during a cluster merger. As a consequence a
bimodality of the CR spatial distribution results with merging (relaxed)
clusters showing a centrally concentrated (flat) CR energy density profile
\cite{article:EnsslinPfrommerMiniatiSubramanian:2011}.  This translates into a bimodality of the expected
diffuse radio and gamma-ray emission of clusters, since more centrally
concentrated CR will find higher target densities for hadronic CR proton
interactions \cite{article:EnsslinPfrommerMiniatiSubramanian:2011}. As a result of this, relaxed clusters
could have a reduced gamma-ray luminosity by up to a factor of five. The goal of
this and other future work of seeking for gamma-ray emission from clusters is to
constrain the acceleration physics and transport properties of CRs.

We adopt the electron density profile for the Coma cluster that has been
inferred from ROSAT X-ray observations \citep{article:BrielHenryBohringer:1992} and model the
spatial and spectral distribution of CRs according to the (conservative) model
by \citet{article:PinzkePfrommer:2010} that neglects the contribution of cluster
galaxies. Figure ?? shows the expected spectrum of Coma, normalized to the total
flux within the virial radius. Add {\em Fermi} results \ldots

@@

To complement the simulation-inspired analysis and to facilitate comparison with
earlier analytical work, we repeat the data analysis with a simplified
analytical model that assumes a power-law CR spectrum and a constant relative CR
pressure, i.e., we adopt the {\em isobaric model of CRs}. Since a priori, the CR
spectral index is unconstrained, we vary it within plausible ranges of
$2.1<\alpha_\rmn{p}<2.5$. These spectral ranges are needed to explain the radio
spectral index of the giant radio halo of Coma after accounting for the spectral
steepening at frequencies $\nu\sim5~\rmn{GHz}$ due to the Sunyaev-Zel'dovich
effect \citep{article:PfrommerEnsslin:2004b}\footnote[4]{Assuming a magnetic field of
  1~$\mu$G, the CR protons responsible for the GHz radio emitting electrons have
  an energy of $~100$~GeV and are $\sim$ 20 times less energetic than those CR
  protons that are responsible for the 200 GeV $\gamma$-ray emission. This is
  consistent with the concave curvature found in the universal CR spectrum by
  \citet{article:PinzkePfrommer:2010}.}.  Table~\ref{table:constraints} shows the
resulting constraints on $X_\CR$ which range between 0.05 and 1, strongly
depending on the assumed spectral index.

@@

%the limit for 0.2 deg and alpha_p = 2.3 is compatible with the simulation based
%one, X_CR < 0.1. However, there is currently a discrepancy in our assumptions
%about the lower energy cutoff of CRs: in the simulations we use P/mp c = 0.8 (as
%simulated) while I set this to 0 in my simplified (isobaric) model. The
%discrepancy can be understood as follows:
%*) in the isobaric model, we would have to normalize our $X_\CR$ by a factor of 2/3
%for an effective spectral index of ~ 2.4 (between GeV and 200 GeV) as inferred
%from the simulations \citep[see Fig 1 in][]{article:PfrommerEnsslin:2004}).
%*) In the simulation model, we have an increased pressure by this factor of 3/2
%(in comparison to a power law spectrum with alpha_TeV ~ 2.3 down to a cutoff in
%the normalized momentum of 0.8) due to the concave curvature of our spectrum.
%}

@@

For clusters that host radio halos we are able to derive a minimum $\gamma$-ray
flux in the hadronic model of CR interactions. The idea is based on the fact
that a stationary distribution of CR electrons loses all its energy to
synchrotron radiation for strong magnetic fields ($B \gg B_\rmn{CMB} \simeq 3.2
\mu$G) so that the ratio of $\gamma$-ray to synchrotron flux becomes independent
of the spatial distribution of CRs and thermal gas \citep{article:Pfrommer:2008}.
This can be easily seen by considering the pion decay induced $\gamma$-ray
luminosity $L_\gamma$ and the synchrotron luminosity $L_\nu$ of a steady state
distribution of CR electrons that has been generated by hadronic CR
interactions,
\begin{eqnarray}
L_\gamma &=& A_\gamma \dps\int d V\, n_\CR n_\rmn{gas},\\
\label{eq:Lnu}
   L_\nu &=& A_\nu \dps\int d V\, n_\CR n_\rmn{gas} \frac{\dps
    \eps_B^{(\alpha_\nu+1)/2}}{\dps\eps_\rmn{CMB}+\eps_B}\\
  &\simeq&A_\nu \dps\int d V\, n_\CR n_\rmn{gas}\quad\rmn{for}~\eps_B \gg \eps_\rmn{CMB}.
\end{eqnarray}
Here $A_\gamma$ and $A_\nu$ are dimensional constants that depend on the
hadronic physics of the interaction \citep{article:PfrommerEnsslinSpringel:2008,
  article:Pfrommer:2008} and $\alpha_\nu\simeq 1$ is the observed synchrotron
spectral index. Hence we can derive a minimum $\gamma$-ray flux in the hadronic
model
\begin{equation}
  \label{eq:Fmin}
  \mathcal{F}_{\gamma,\rmn{min}} = 
  \frac{\dps A_\gamma}{\dps A_\nu}\frac{\dps L_\nu}{\dps 4\pi D_\rmn{lum}^2},
\end{equation}
where $L_\nu$ is the observed luminosity of the radio mini-halo and
$D_\rmn{lum}$ denotes the luminosity distance to the respective cluster.
Lowering the magnetic field would require an increase in the energy density of
CR electrons to reproduce the observed synchrotron luminosity and thus increase
the associated $\gamma$-ray flux.

To derive a minimum gamma-ray flux that is comparable to the upper limits, we
need to determined the radio flux within the corresponding angular regions. To
this end, we fit the point source subtracted, azimuthally averaged radio halo
profile at 1.38 GHz \citep{article:Deiss_etal:1997} with a $\beta$-model,
\begin{equation}
S_\nu (r_\bot)= S_0\, 
\left[ 1 + \left( \frac{r_\bot}{r_{\rmn{c}}}\right)^2\right]
^{-3\beta + 1/2},
\label{beta}
\end{equation}
where $S_0 = 1.1 \times 10^{-3}\,\rmn{Jy\,arcmin}^{-2}$, $r_{\rmn{c}} = 450
kpc$, and $\beta = 0.78$.  Within the error bars, this profile is consistent
with 326 MHz-data taken by \citet{article:Govoni_etal:2001} when scaled with a radio
spectral index of 1.15.

@@

We have seen in the previous Section that we can obtain an absolute lower limit
on the gamma-ray emission in the hadronic model by assuming high magnetic
fields, $B>B_\rmn{CMB}$. Hence, we can turn this around by using our upper limit
on the gamma-ray emission (and by extension on the CR pressure) to infer a lower
limit on the magnetic field that is needed to explain the observed radio
emission which, however, assumes the validity of the hadronic model. Lowering
the gamma-ray constraint will tighten (increase) the magnetic field limit. Once
this gets in conflict with magnetic field measurements by means of other
methods, e.g., Faraday rotation measure (RM)\footnote{Generally, Faraday RM
  of the magnetic field strength by, e.g., background sources through clusters
  are degenerate with the magnetic coherence scale. Only a simultaneous
  deprojection method that self-consistently solves for the magnetic power
  spectrum is in principle able to break this degeneracy, however, require
  high-quality data and picks up further complications when using embedded
  sources within the cluster plasma, e.g., degeneracies with the assumed
  projection angle of the jet \citep{article:KucharEnsslin:2011}.}, this would
challenge the hadronic model of radio halos.  The method that we use to
constrain the magnetic field inherits a dependence on the assumed spatial
structure that we parametrize by
\begin{equation}
  \label{eq:B}
  B(r) = B_0  \,\left(\frac{n_\rmn{e}(r)}{n_\rmn{e}(0)}\right)^{\alpha_B},
\end{equation}
as suggested by Faraday RM studies and numerical magneto-hydrodynamical (MHD)
simulations \citep[][and references therein]{article:Bonafede_etal:2010, article:Bonafede_etal:2011}. In fact, the magnetic field in the Coma cluster is among
the best ones constrained due to its proximity that enables RM observations of
seven radio-sources located at projected distances of 50 to 1500 kpc from the
cluster center. The best fit model obtains $B_0 = 4.7^{+0.7}_{-0.8}\,\mu$G and
$\alpha_B = 0.5^{+0.2}_{-0.1}$ \citep{article:Bonafede_etal:2010}.  We aim for
constraining the central field strength, $B_0$, which depends on the magnetic
decline, $\alpha_B$, that we vary (within reasonable ranges of
$\Delta\alpha_B=0.2$ as suggested from the Faraday RM studies) to address this
dependence. We proceed as follows:
